Global search algorithm with local tuning for solving mixed-integer global optimization problems \thanks{This study was supported by the .}

### 2021-04-08 2021/MOTOR_BarkalovLebedev/MOTOR2021.tex
MOTOR правки

\begin{document}
%
\title{An Approach for Simultaneous Finding of Multiple Efficient Decisions in Multi-objective Optimization Problems
%with non-convex constraints 
\thanks{This work was supported by the Ministry of Science and Higher Education of the Russian Federation, project no.0729-2020-0055, and by the Research and Education Mathematical Center, project no. 075-02-2020-1483/1.}}
%
\titlerunning{Finding of Multiple Effective Decisions in MOO Problems}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Konstantin Barkalov\orcidID{0000-0001-5273-2471} \and
Victor Gergel\orcidID{0000-0002-4013-2329} \and
Vladimir Grishagin\orcidID{0000-0002-2884-3670} \and
Evgeniy Kozinov\orcidID{0000-0001-6776-0096} 
}
%
\authorrunning{K. Barkalov, V. Gergel, V. Grishagin, E. Kozinov}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Lobachevsky State University of Nizhni Novgorod, Nizhni Novgorod, Russia 
\email{\{konstantin.barkalov,evgeny.kozinov\}@itmm.unn.ru,\{gergel,vagris\}@unn.ru}
}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
This paper considers computationally intensive multi-objective optimization problems which require computing multiple Pareto-optimal decisions. It is also assumed that efficiency criteria may be multiextremal, and the cost of calculating function values may be quite high. The proposed approach is based on the reduction of multi-objective optimization problems to one-dimensional global optimization problems that can be solved using efficient information-statistical algorithms of global search. One key innovation of the developed approach consists in the possibility of solving several global optimization problems simultaneously, which allows multiple Pareto-optimal decisions to be obtained. Besides, such approach provides for reuse of the computed search information, which considerably reduces computational effort for solving multi-objective optimization problems. Computational experiments confirm the potential of the proposed approach.

@@

Choosing the optimal options (decision making) in situations with many different alternatives is a problem that occurs in almost every domain of human activity. In many cases, decision making problems can be viewed as optimization problems of various kinds such as convex programming, discrete optimization, nonlinear programming, etc. In more complex situations, the objective functions that determine the effectiveness of decisions can be multiextremal, and decision making will require solving global optimization problems. In the most general case, however, the effectiveness of decisions may be determined by several objective criteria, which necessitates solving multi-objective optimization (MOO) problems. Thus, MOO problems are the most common decision making statements that one has to solve in many scientific and technical applications. A large number of approaches and methods have been developed to solve such problems and they have been used to solve many decision making problems in various fields of practical applications -- see e.g. \cite{c1,c2,c3,c4,c5,c6,c7,c8,c9}.

@@

It is also important to note that by solving simultaneously a large number of scalar optimization problems thus generated it is possible to significantly decrease computational complexity of each separate problem. This effect is achived due to the fact that all such scalar problems  are based on the same MOO problem and, consequently, all computed values of the efficiency criteria of the MOO problem can be reduced to values of any scalar problem being solved simultaneously without any time consuming calculations. In such cases, all the search information obtained in solving any single scalar problem can be used for solving all other scalar problems of the same group.

This paper presents the results of the authors continuing research aimed to develop effective methods for solving MOO problems \cite{c35,c36,c37,c44}. In \cite{c35}, some methods for solving global optimization problems are considered, in \cite{c36},  a general decision-making model based on MOO problems is proposed, and in \cite{c37}, a theoretical analysis is presented regarding the effectiveness of the use of search information when solving MOO problems. In \cite{c44}, we consider MOO problems in which the efficiency criteria can be ordered by importance (lexicographic multi-objective optimization). In this case, MOO problems are reduced to global optimization problems with non-convex constraints for which constrained optimization methods should be applied. In this paper, we propose new computational schemes for using search information that significantly reduce the amount of calculations required to solve MOO problems.

The structure of the paper is as follows. In Section \ref{sec:02}, we give the statement of the multi-objective optimization problem. Section \ref{sec:03} presents a scheme for reduction of multi-objective optimization problems to one-dimensional global optimization problems as well as some methods for solving such problems. Section \ref{sec:04} considers the proposed approach for simultaneous solution of several global optimization problems yielding several Pareto-optimal decisions at a time. Section \ref{sec:05} contains the results of numerical experiments confirming the effectiveness of the proposed approach. In conclusion, the results obtained are discussed and possible main lines of action for further research are outlined.

@@

\subsection{Scalarization of multiple objective functions} 
In the most general form, the global optimization problem generated by the scalarization of multiple objective functions of the MOO problem can be represented in the form
\begin{equation}
\label{eq:04}
\min_{y \in D} {\varphi(y)} = \min_{y \in D} F(\lambda, y),
\end{equation}
where $F$ is a scalar multiextremal function generated as a result of scalarization of objective functions $f_i$, $1 \leq i \leq s$, $\lambda$ is the vector of parameters of the applied convolution of functions, and $D$ is the search domain from (\ref{eq:01}). In the proposed approach, we use for scalarization the compromise programming method \cite{c2,c11}, where solution of the MOO problem consists in finding the efficient decision corresponding most closely to the values of the objective functions of the specified reference decision $y^0 \in D$. In this case, a possible statement of scalar optimization problem can have the form 
\begin{equation}
\label{eq:05}
\min_{y \in D} F(\lambda,y)=\min_{y \in D} \left\{ \frac{1}{s} \sum_{i=1}^s {\lambda_i (f_i (y)-f_i (y^0 ))^2} \right\},
\end{equation}
where $F(\lambda,y)$ is the standard deviation of the values of the objective functions $f_i$, $1 \leq i \leq s$, for the decision $y \in D$ and for the specified reference decision $y^0 \in D$, while the coefficients $\lambda_i$, $0 \leq \lambda_i \leq 1$, $1 \leq i \leq s$, are measures of the significance of the approximation accuracy for each variable $y_i$, $1 \leq i \leq N$, separately. Without loss of generality, we can assume that the area of possible values of the coefficients $\lambda$ is a set 
\begin{equation}
\label{eq:06}
\lambda=(\lambda_1,\lambda_2, \dots, \lambda_s)\in \Lambda \subset R^s: \sum_{i=1}^{s}\lambda_i=1, \lambda_i \geq 0, 1 \leq i \leq s.
\end{equation}

@@

\subsection{Dimensionality reduction} 
As already mentioned, problem (\ref{eq:04}) is a multidimensional global optimization problem. Problems of this kind are computationally complex and are known to be subject to the ``curse of dimensionality'' -- computational complexity increases exponentially with increasing dimensionality of the optimization problem being solved \cite{c13,c14,c15,c16,c17,c18,c19,c20}. Nevertheless, the computational complexity of global optimization algorithms can be significantly reduced by dimensionality reduction based on the use of a Peano space-filling curve (or evolvent) $y(x)$ that uniquely and continuously maps the segment $[0,1]$ on an $N$-dimensional hypercube $D$ -- see, for example, \cite{c15,c21}. As a result of such reduction, multidimensional global optimization problems (\ref{eq:04}) are reduced to one-dimensional problems
\begin{equation}
\label{eq:09}
\min_{x \in [0,1]} \varphi (y(x))= \min_{y \in D} \varphi(y).
\end{equation}

The resulting one-dimensional functions $\varphi(y(x))$ satisfy the uniform H{\" o}lder condition, i.e.
\begin{equation}
\label{eq:10}
|\varphi (y(x'))- \varphi (y(x''))| \leq H |x'-x''|^{1/N}, \; x',x''\in[0,1],
\end{equation}
where the constant $H$ is defined by the relation $H=2L\sqrt{N+3}$, $L$ is the Lipschitz constant from (\ref{eq:08}), and $N$ is the dimensionality of the optimization problem (\ref{eq:01}).

\subsection{Solving the one-dimensional reduced optimization problem} 
Using the dimensionality reduction results in one more additional advantage of the proposed approach: many well-known one-dimensional global search algorithms (possibly, after some additional generalisation) can be used to solve the initial multidimensional MOO problem from (\ref{eq:01}) \cite{c22,c23,c24,c25,c26,c27,c28,c29}. At the same time, it should be noted that most works where dimensionality reduction is a key feature for solving multiextremal problems like (\ref{eq:04}) rely on the information-statistical theory of global search \cite{c15}. This theory has provided the a basis for developing a large number of efficient methods for multi-extremal optimization \cite{c28,c29,c30,c31,c32,c33,c34,c35,c36,c37}.

@@

For the algorithms proposed in \cite{c22,c23}, when dimensionality reduction is applied, the characteristic is an estimate of the minimum possible value of the function $\varphi(y(x))$ to be minimized on the interval $(x_{i-1},x_i )$, $1 < i \leq k$, i.e.
\begin{equation}
\label{eq:13}
R(i) = 0.5 H \rho_i - 0.5 (z_{i-1} + z_i ), \rho_i = (x_i-x_{i-1} ) ^ {1⁄N}, 1 < i \leq k,
\end{equation}
where $H$ is the H{\" o}lder constant from (\ref{eq:10}) for the reduced global optimization problem being solved (\ref{eq:09}), $z_i = \varphi(y(z_i))$, $1 \leq i \leq k,$ and $N$ is the dimensionality of the problem from (\ref{eq:01})).  For the global search algorithm (GSA) \cite{c15,c24} developed in the framework of the information-statistical approach, the characteristic is
\begin{equation}
\label{eq:14}
R(i)=m\rho_i+\frac{z_i-z_{i-1}^2}{m\rho_i}-2(z_{i-1}+z_i ),\rho_i=(x_i-x_{i-1})^{1/N}, 1 < i \leq k,
\end{equation}
where $m$ is a numerical estimate of the H{\" o}lder constant derived from available search information
\begin{equation}
\label{eq:15}
m= r M, M = \max \{ |z_i-z_{i-1} | ⁄ \rho_i, 1 < i \leq  k \}
\end{equation}
 ($r>1$ is a parameter of the GSA algorithm).

The presence of interval characteristics makes it possible to present the procedure of global search iteration as the following sequence of steps \cite{c15}.

@@

After completing the calculations, the lowest computed value of the function being minimized can be taken as the global minimum estimate
\begin{equation}
\label{eq:19}
z_k^* = \min \{ z_i, 1 \leq i \leq \}.
\end{equation}

@@

Convergence conditions of the algorithms formulated in the form (\ref{eq:16})--(\ref{eq:18}). One of sufficient conditions for convergence of algorithms is, for example, the requirement that the characteristic of the interval containing the global minimum point should take on the maximum value at step 1 of the scheme (\ref{eq:16})--(\ref{eq:18}) during the global search iterations. This condition is satisfied, for example, for the multidimensional generalized algorithms proposed in \cite{c22,c23} when the H{\"o}lder constant from (\ref{eq:10}) is specified exactly. For GSA, a sufficient condition for convergence is the relation \cite{c15}
\begin{equation}
\label{eq:20}
m \geq 2^{3-1⁄N} L\sqrt{N+3},
\end{equation}
which must be fulfilled starting from some iteration $k>1$ of the global search ($L$ is the Lipschitz constant from (\ref{eq:08})). Moreover, if condition (\ref{eq:20}) is satisfied, only the points of the global minimum of the function $\varphi(y)$ from (\ref{eq:04}) will be the limit points of the trial sequence $\{y^k=y(x^k)\}$ generated by the GSA algorithm.

@@

As mentioned earlier, in the process of solving the MOO problem, it may be necessary to find several different efficient decisions due to possible changes in the optimality requirements. Obtaining different efficient decisions in the proposed approach is ensured by choosing different convolution coefficients (importance indicators) for the objective functions $f_i(y)$, $1 \leq i \leq s$, which results in obtaining different  scalar multi-extremal functions $F(\lambda, y)$ from (\ref{eq:04}). These functions $F(\lambda, y)$ can be optimized sequentially. This determines a multi-stage scheme for solving the MOO problem, when at each stage DM specifies the necessary scalarization coefficients, then the resulting scalar optimization problem is solved, after which DM analyzes the efficient decisions found.

The general scheme considered above can be extended by the option of selecting not one but several different scalarization coefficients at each stage. Having such an option makes it easier for DM  to specify importance coefficients $\lambda \in \Lambda$ of the objective functions $f_i(y)$, $1 \leq i \leq s$. By solving simultaneously several generated scalar optimization problems, one can  obtain the estimates of efficient decisions at the earliest stages of computing, which makes it possible to dynamically (in the process of computations) change the set of problems being solved -- to stop solving those that obviously have no prospect of success (from DM's point of view) or to add new optimization problems.

Such generalization of the process of solving a MOO problem means that at each current moment of calculations there is a set of functions being optimized simultaneously having the form (\ref{eq:04})
\begin{equation}
\label{eq:21}
\Phi_p (y)=\{ F(\lambda_1,y), F(\lambda_2,y),\dots,F(\lambda_p,y)  \},  \lambda_i \in \Lambda, 1 \leq i \leq p,
\end{equation}
which can be changed dynamically in the course of the calculations by adding new or removing existing optimization functions $F(\lambda, y)$ from (\ref{eq:04}).

@@

As shown in Section \ref{sec:03}, the information-statistical multiextremal optimization algorithms used in the proposed approach determine the points of consecutive iterations of the global search taking into account the search information
\begin{equation}
\label{eq:22}
A_k=\{(x_i,z_i,f_i )^T:1 \leq i \leq k\}
\end{equation}
obtained in the calculations (see (\ref{eq:14})--(\ref{eq:17})). In (\ref{eq:22}) $x_i$, $1 \leq i \leq k$,  are the reduced points of performed global search iterations ordered in ascending order of coordinates,  $z_i$, $f_i$, $1 \leq i \leq k$, are values of the scalar function $F(\lambda, y)$, from (\ref{eq:04}) and objective functions $f(y)$ from (\ref{eq:01}) of the current optimization problem to be solved at the points $x_i=y(x_i)$, $1 \leq i \leq k$. By using search information $A_k$ from (\ref{eq:22}) when choosing next search iterations it is possible to solve global optimization problems more efficiently and to provide convergence of algorithms only to the global minima of multiextremal functions being minimized.

It is important to note that since the set $\Phi_p(y)$ of functions being optimized  simultaneously is generated from the same MOO problem from (\ref{eq:01}), the existence of the set $A_k$ from (\ref{eq:22}) allows us to adjust the results of all previously performed calculations of the values of the objective functions $f(y)$ to the values of the next optimized function $F(\lambda, y)$ from (\ref{eq:04}) without repeating any time consuming calculations of values, i.e.
\begin{equation}
\label{eq:23}
(x_i,f_i ) \to z_i=F(\lambda, y(x_i)) ,1 \leq i \leq k.
\end{equation}

@@

Accordingly, for each interval $(x_{i-1},x_i)$, $1 < i \leq k$, into which the segment [0,1] is divided, the following set of characteristics will be calculated:
\begin{equation}
\label{eq:26}
\overrightarrow{R}(i)=\{R_1(i), R_2(i), \dots, R_p(i)\},
\end{equation}
where
\begin{equation}
\label{eq:27}
R_j(i)=m_j \rho_i+\frac{(z_i(j)-z_{i-1}(j))^2}{m_j \rho_i} - 2(z_{i-1}(j)+z_i(j)),1 < i \leq k,1 \leq j \leq p,
\end{equation}
\begin{equation}
\label{eq:28}
\rho_i=(x_i-x_{i-1} )^{1/N}  ,1 \leq i<k,
\end{equation}
\begin{equation}
\label{eq:29}
m_j = r M_j, M_j=\left\{ \frac{\max|z_i (j)-z_{i-1} (j)|}{\rho_i} ,1< i \leq k, 1 \leq j \leq \right\}.
\end{equation}
($m_j$, $1 \leq j \leq p$, are the estimates of the H{\"o}lder  constant in the condition (\ref{eq:10}) for the functions $F(\lambda_j,y)$, $1 \leq j \leq p$, of the set $\Phi_p(y)$ from (\ref{eq:21})).

@@

Numerical experiments were performed on the Lobachevsky supercomputer of the University of Nizhni Novgorod (operating system -- CentOS 6.4, control system -- SLURM). One supercomputer node has 2 Intel Sandy Bridge E5-2660 2.2 GHz processors, 64 Gb RAM. The CPU is 8-core (i.e. a total of 16 CPU cores are available on the node). The numerical experiments were performed using the Globalizer system \cite{c38}.

@@

For MGSA 100 problems (\ref{eq:04}) with different values of convolution coefficients $\lambda$ uniformly distributed in $\Lambda$ were solved. The full results of the experiments performed are presented in Table \ref{tab:1}.

@@

The results of numerical experiments are presented in Table \ref{tab:2} and fig. \ref{fig:1}. The table shows in the first column the number of functions $\Phi_p (y)$ from (\ref{eq:21}) being optimized.  The columns named Iters contain the average number of iterations performed by the algorithm to solve the MOO problem. The columns named HV and DU contain the values of the HV and DU indicators. The results of the numerical experiments are divided with respect to the three algorithms used: GSA is described in Section \ref{sec:03}, MGSA is discussed in Section \ref{sec:03} (item 2), SGSA is presented in Section \ref{sec:04} (item 3).

@@

In the third series of numerical experiments, 10 bi-criteria five-dimensional MOO problems were solved, i.e. $N = 5$, $s = 2$, $m=2$. The objective functions of the MOO problems were determined, as before, using the GKLS generator \cite{c43}. When solving the problems, the accuracy $\varepsilon=0.05$ and the reliability $r=5.6$ were used. The results of numerical experiments are presented in Table \ref{tab:4} and fig. \ref{fig:2}.

@@

This paper proposes an efficient method for solving computationally intensive multi-objective optimization problems which require computing several Pareto-optimal decisions. It is also assumed that objective functions may be multiextremal, and computation of function values may require a large amount of calculations. The proposed approach is based on the reduction of multi-objective optimization problems to one-dimensional global optimization problems that can be solved using efficient information-statistical algorithms of global search.  The novelty of the proposed approach consists in the possibility of solving several global optimization problems simultaneously, which allows multiple Pareto-optimal decisions to be obtained. Besides, such approach provides for reuse of the computed search information, which considerably reduces computational effort for solving multi-objective optimization problems.

>>>>

\begin{document}
%
\title{An Approach for Simultaneous Finding of Multiple Efficient Decisions in Multi-objective Optimization Problems
%with non-convex constraints 
\thanks{This work was supported by the Ministry of Science and Higher Education of the Russian Federation, project no.0729-2020-0055, and by the Research and Education Mathematical Center, project no. 075-02-2020-1483/1.}}
%
\titlerunning{Finding of Multiple Efficient Decisions in MOO Problems}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Konstantin Barkalov\orcidID{0000-0001-5273-2471} \and
Victor Gergel\orcidID{0000-0002-4013-2329} \and
Vladimir Grishagin\orcidID{0000-0002-2884-3670} \and
Evgeniy Kozinov\orcidID{0000-0001-6776-0096} 
}
%
\authorrunning{K. Barkalov, V. Gergel, V. Grishagin, E. Kozinov}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Lobachevsky State University of Nizhni Novgorod, Nizhni Novgorod, Russia 
\email{\{konstantin.barkalov,evgeny.kozinov\}@itmm.unn.ru,\{gergel,vagris\}@unn.ru}
}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
This paper considers computationally intensive multi-objective optimization problems which require computing multiple Pareto-optimal decisions. It is also assumed that efficiency criteria may be multiextremal, and the cost of calculating function values may be quite high. The proposed approach is based on the reduction of multi-objective optimization problems to one-dimensional global optimization problems that can be solved using efficient information-statistical algorithms of global search. One key innovation of the developed approach consists in the possibility of solving several global optimization problems simultaneously, which allows multiple Pareto-optimal decisions to be obtained. Besides, such approach provides for reuse of the computed search information, which considerably reduces computational effort for solving multi-objective optimization problems. Computational experiments confirm the potential of the proposed approach.

@@

Choosing the optimal decisions (decision making) in situations with many different alternatives is a problem that occurs in almost every domain of human activity. In many cases, decision making problems can be viewed as optimization problems of various kinds such as convex programming, discrete optimization, nonlinear programming, etc. In more complex situations, the objective functions that determine the effectiveness of decisions can be multiextremal, and decision making will require solving global optimization problems. In the most general case, however, the effectiveness of decisions may be determined by several objective criteria, which necessitates solving multi-objective optimization (MOO) problems. Thus, MOO problems are the most common decision making statements that one has to solve in many scientific and technical applications. A large number of approaches and methods have been developed to solve such problems and they have been used to solve many decision making problems in various fields of practical applications -- see e.g. \cite{c1,c2,c3,c4,c5,c6,c7,c8,c9}.

@@

It is also important to note that by solving simultaneously a large number of scalar optimization problems thus generated it is possible to significantly decrease computational complexity of each separate problem. This effect is achived due to the fact that all such scalar problems  are based on the same MOO problem and, consequently, all computed values of the efficiency criteria of the MOO problem can be reduced to values of any scalar problem being solved simultaneously without any time consuming calculations. In such cases, all the search information obtained in solving any single scalar problem can be used for solving all other scalar problems of the same set.

This paper presents the results of the authors continuing research aimed to develop efficient methods for solving MOO problems \cite{c35,c36,c37,c44}. In \cite{c35}, some methods for solving global optimization problems are considered, in \cite{c36},  a general decision-making model based on MOO problems is proposed, and in \cite{c37}, a theoretical analysis is presented regarding the effectiveness of the use of search information when solving MOO problems. In \cite{c44}, we consider MOO problems in which the efficiency criteria can be ordered by importance (lexicographic multi-objective optimization). In this case, MOO problems are reduced to global optimization problems with non-convex constraints for which constrained optimization methods should be applied. In this paper, we propose new computational schemes for using search information that significantly reduce the amount of calculations required to solve MOO problems.

The structure of the paper is as follows. In Section \ref{sec:02}, we give the statement of the multi-objective optimization problem. Section \ref{sec:03} presents a scheme for reduction of multi-objective optimization problems to one-dimensional global optimization problems as well as some methods for solving such problems. Section \ref{sec:04} considers the proposed approach for simultaneous solution of several global optimization problems yielding several Pareto-optimal decisions at a time. Section \ref{sec:05} contains the results of numerical experiments confirming the effectiveness of the proposed approach. In conclusion, the results obtained are discussed and possible main directions of further research are outlined.

@@

\subsection{Scalarization of multiple objective functions}
\label{subsec:31}

In the most general form, the global optimization problem generated by the scalarization of multiple objective functions of the MOO problem can be represented in the form
\begin{equation}
\label{eq:04}
\min_{y \in D} {\varphi(y)} = \min_{y \in D} F(\lambda, y),
\end{equation}
where $F$ is a scalar multiextremal function generated as a result of scalarization of objective functions $f_i$, $1 \leq i \leq s$, $\lambda$ is the vector of parameters of the applied convolution of functions, and $D$ is the search domain from (\ref{eq:01}). In the proposed approach, we use for scalarization the compromise programming method \cite{c2,c11}, where solution of the MOO problem consists in finding the efficient decision corresponding most closely to the values of the objective functions of the specified reference decision $y^0 \in D$. In this case, a possible statement of scalar optimization problem can have the form 
\begin{equation}
\label{eq:05}
\min_{y \in D} F(\lambda,y)=\min_{y \in D} \left\{ \frac{1}{s} \sum_{i=1}^s {\lambda_i (f_i (y)-f_i (y^0 ))^2} \right\},
\end{equation}
where $F(\lambda,y)$ is the standard deviation of the values of the objective functions $f_i$, $1 \leq i \leq s$, for the decision $y \in D$ and for the specified reference decision $y^0 \in D$, while the coefficients $\lambda_i$, $0 \leq \lambda_i \leq 1$, $1 \leq i \leq s$, are measures of the importance of the approximation accuracy for each variable $y_i$, $1 \leq i \leq N$, separately. Without loss of generality, we can assume that the domain of possible values of the coefficients $\lambda$ is a set 
\begin{equation}
\label{eq:06}
\lambda=(\lambda_1,\lambda_2, \dots, \lambda_s)\in \Lambda \subset R^s: \sum_{i=1}^{s}\lambda_i=1, \lambda_i \geq 0, 1 \leq i \leq s.
\end{equation}

@@

\subsection{Dimensionality reduction} 
\label{subsec:32}

As already mentioned, problem (\ref{eq:04}) is a multidimensional global optimization problem. Problems of this kind are computationally complex and are known to be subject to the ``curse of dimensionality'' -- computational complexity increases exponentially with increasing dimensionality of the optimization problem being solved \cite{c13,c14,c15,c16,c17,c18,c19,c20}. Nevertheless, the computational complexity of global optimization algorithms can be significantly reduced by dimensionality reduction based on the use of a Peano space-filling curve (or evolvent) $y(x)$ that uniquely and continuously maps the segment $[0,1]$ on an $N$-dimensional domain $D$ -- see, for example, \cite{c15,c21}. As a result of such reduction, multidimensional global optimization problems (\ref{eq:04}) are reduced to one-dimensional problems
\begin{equation}
\label{eq:09}
\min_{x \in [0,1]} \varphi (y(x))= \min_{y \in D} \varphi(y).
\end{equation}

The resulting one-dimensional functions $\varphi(y(x))$ satisfy the uniform H{\" o}lder condition, i.e.
\begin{equation}
\label{eq:10}
|\varphi (y(x'))- \varphi (y(x''))| \leq H |x'-x''|^{1/N}, \; x',x''\in[0,1],
\end{equation}
where the constant $H$ is defined by the relation $H=2L\sqrt{N+3}$, $L$ is the Lipschitz constant from (\ref{eq:08}), and $N$ is the dimensionality of the MOO problem (\ref{eq:01}).

\subsection{Solving the one-dimensional reduced optimization problem} 
\label{subsec:33}

Using the dimensionality reduction results in one more additional advantage of the proposed approach: many well-known one-dimensional global search algorithms (possibly, after some additional generalisation) can be used to solve the initial multidimensional MOO problem from (\ref{eq:01}) \cite{c22,c23,c24,c25,c26,c27,c28,c29}. At the same time, it should be noted that most works where dimensionality reduction is a key feature for solving multiextremal problems like (\ref{eq:04}) rely on the information-statistical theory of global search \cite{c15}. This theory has provided the a basis for developing a large number of efficient methods for multiextremal optimization \cite{c28,c29,c30,c31,c32,c33,c34,c35,c36,c37}.

@@

For the algorithms proposed in \cite{c22,c23}, when dimensionality reduction is applied, the characteristic is an estimate of the minimum possible value of the function $\varphi(y(x))$ to be minimized on the interval $(x_{i-1},x_i )$, $1 < i \leq k$, i.e.
\begin{equation}
\label{eq:13}
R(i) = 0.5 H \rho_i - 0.5 (z_{i-1} + z_i ), \rho_i = (x_i-x_{i-1} )^{1⁄N}, 1 < i \leq k,
\end{equation}
where $H$ is the H{\" o}lder constant from (\ref{eq:10}) for the reduced global optimization problem being solved (\ref{eq:09}), $z_i = \varphi(y(z_i))$, $1 \leq i \leq k,$ and $N$ is the dimensionality of the problem from (\ref{eq:01})).  For the global search algorithm (GSA) \cite{c15,c24} developed in the framework of the information-statistical approach, the characteristic is
\begin{equation}
\label{eq:14}
R(i)=m\rho_i+\frac{(z_i-z_{i-1})^2}{m\rho_i}-2(z_{i-1}+z_i ),\rho_i=(x_i-x_{i-1})^{1/N}, 1 < i \leq k,
\end{equation}
where $m$ is a numerical estimate of the H{\" o}lder constant derived from available search information
\begin{equation}
\label{eq:15}
m= r M, M = \max \{ |z_i-z_{i-1} | ⁄ \rho_i, 1 < i \leq  k \}
\end{equation}
 ($r>1$ is a parameter of the GSA algorithm).

The presence of interval characteristics makes it possible to describe the procedure of global search iteration as the following sequence of steps \cite{c15}.

@@

After completing the calculations, the lowest computed value of the function being minimized can be taken as the global minimum estimate
\begin{equation}
\label{eq:19}
z_k^* = \min \{ z_i, 1 \leq i \leq k\}.
\end{equation}

@@

Convergence conditions of the algorithms formulated in the form (\ref{eq:16})--(\ref{eq:18}) depend on the properties of the interval characteristics used. One of sufficient conditions for convergence of algorithms is, for example, the requirement that the characteristic of the interval containing the global minimum point should take on the maximum value at step 1 of the scheme (\ref{eq:16})--(\ref{eq:18}) during the global search iterations. This condition is satisfied, for example, for the multidimensional generalized algorithms proposed in \cite{c22,c23} when the H{\"o}lder constant from (\ref{eq:10}) is specified exactly. For GSA, a sufficient condition for convergence is the relation 
\begin{equation}
\label{eq:20}
m \geq 2^{3-1⁄N} L\sqrt{N+3},
\end{equation}
which must be fulfilled starting from some iteration $k>1$ of the global search ($L$ is the Lipschitz constant from (\ref{eq:08})) \cite{c15}. Moreover, if condition (\ref{eq:20}) is satisfied, only the points of the global minimum of the function $\varphi(y)$ from (\ref{eq:04}) will be the limit points of the trial sequence $\{y^k=y(x^k)\}$ generated by the GSA algorithm.

@@

As mentioned earlier, in the process of solving the MOO problem, it may be necessary to find several different efficient decisions due to possible changes in the optimality requirements. Obtaining different efficient decisions in the proposed approach is ensured by choosing different convolution coefficients (importance indicators) for the objective functions $f_i(y)$, $1 \leq i \leq s$, which results in obtaining different  scalar multiextremal functions $F(\lambda, y)$ from (\ref{eq:04}). These functions $F(\lambda, y)$ can be optimized sequentially. This determines a multi-stage scheme for solving the MOO problem, when at each stage DM specifies the necessary scalarization coefficients, then the resulting scalar optimization problem is solved, after which DM analyzes the efficient decisions found.

The general scheme considered above can be extended by the possibility of selecting not one but several different scalarization coefficients at each stage. Having such an option makes it easier for DM  to specify importance coefficients $\lambda \in \Lambda$ of the objective functions $f_i(y)$, $1 \leq i \leq s$. By solving simultaneously several generated scalar optimization problems, one can  obtain the estimates of efficient decisions at the earliest stages of computing, which makes it possible to dynamically (in the process of computations) change the set of problems being solved -- to stop solving those that obviously have no prospect of success (from DM's point of view) or to add new optimization problems.

Such generalization of the process of solving a MOO problem means that at each current moment of calculations there is a set of functions being optimized simultaneously having the form
\begin{equation}
\label{eq:21}
\Phi_p (y)=\{ F(\lambda_1,y), F(\lambda_2,y),\dots,F(\lambda_p,y)  \},  \lambda_i \in \Lambda, 1 \leq i \leq p,
\end{equation}
which can be changed dynamically in the course of the calculations by adding new or removing existing optimization functions $F(\lambda, y)$ from (\ref{eq:04}).

@@

As shown in Section \ref{sec:03}, the information-statistical multiextremal optimization algorithms used in the proposed approach determine the points of consecutive iterations of the global search taking into account the search information
\begin{equation}
\label{eq:22}
A_k=\{(x_i,z_i,f_i )^T:1 \leq i \leq k\}
\end{equation}
obtained in the calculations (see (\ref{eq:14})--(\ref{eq:17})). In (\ref{eq:22}) $x_i$, $1 \leq i \leq k$,  are the reduced points of performed global search iterations ordered in ascending order of coordinates,  $z_i$, $f_i$, $1 \leq i \leq k$, are values of the scalar function $F(\lambda, y)$, from (\ref{eq:04}) and objective functions $f_i(y)$, $1 \leq i \leq s$, from (\ref{eq:01}) of the current optimization problem to be solved at the points $x_i=y(x_i)$, $1 \leq i \leq k$. By using search information $A_k$ from (\ref{eq:22}) when choosing next search iterations it is possible to solve global optimization problems more efficiently and to provide convergence of algorithms only to the global minima of multiextremal functions being minimized.

It is important to note that since the set $\Phi_p(y)$ of functions being optimized  simultaneously is generated from the same MOO problem from (\ref{eq:01}), the existence of the set $A_k$ from (\ref{eq:22}) allows us to adjust the results of all previously performed calculations of the values of the objective functions $f_i(y)$, $1 \leq i \leq s$, to the values of the next optimized function $F(\lambda, y)$ from (\ref{eq:04}) without repeating any time consuming calculations of values, i.e.
\begin{equation}
\label{eq:23}
(x_i,f_i ) \to z_i=F(\lambda, y(x_i)) ,1 \leq i \leq k.
\end{equation}

@@

Accordingly, for each interval $(x_{i-1},x_i)$, $1 < i \leq k$, into which the segment [0,1] is divided, the following set of characteristics will be calculated:
\begin{equation}
\label{eq:26}
\overrightarrow{R}(i)=\{R_1(i), R_2(i), \dots, R_p(i)\},
\end{equation}
where
\begin{equation}
\label{eq:27}
R_j(i)=m_j \rho_i+\frac{(z_i(j)-z_{i-1}(j))^2}{m_j \rho_i} - 2(z_{i-1}(j)+z_i(j)),1 < i \leq k,1 \leq j \leq p,
\end{equation}
\begin{equation}
\label{eq:28}
\rho_i=(x_i-x_{i-1} )^{1/N}  ,1 \leq i<k,
\end{equation}
\begin{equation}
\label{eq:29}
m_j = r M_j, M_j=\left\{ \frac{\max|z_i (j)-z_{i-1} (j)|}{\rho_i} ,1< i \leq k, 1 \leq j \leq p \right\}.
\end{equation}
($m_j$, $1 \leq j \leq p$, are the estimates of the H{\"o}lder  constant in the condition (\ref{eq:10}) for the functions $F(\lambda_j,y)$, $1 \leq j \leq p$, of the set $\Phi_p(y)$ from (\ref{eq:21})).

@@

Numerical experiments were performed on the Lobachevsky supercomputer of the University of Nizhni Novgorod (operating system -- CentOS 6.4, management system -- SLURM). One supercomputer node has 2 Intel Sandy Bridge E5-2660 2.2 GHz processors, 64 Gb RAM. The CPU is 8-core (i.e. a total of 16 CPU cores are available on the node). The numerical experiments were performed using the Globalizer system \cite{c38}.

@@

For MGSA 100 problems (\ref{eq:04}) with different values of convolution coefficients $\lambda$ uniformly distributed in $\Lambda$ were solved. The results of the experiments performed are presented in Table \ref{tab:1}.

@@

The results of numerical experiments are presented in Table \ref{tab:2} and Fig. \ref{fig:1}. The Table shows in the first column the number of functions $\Phi_p (y)$ from (\ref{eq:21}) being optimized.  The columns named Iters contain the average number of iterations performed by the algorithm to solve the MOO problem. The columns named HV and DU contain the values of the HV and DU indicators. The results of the numerical experiments are divided with respect to the three algorithms used: GSA is described in Section \ref{sec:03}, MGSA is discussed in Section \ref{subsec:32}, SGSA is presented in Section \ref{subsec:43}.

@@

In the third series of numerical experiments, 10 bi-criteria five-dimensional MOO problems were solved, i.e. $N = 5$, $s = 2$. The objective functions of the MOO problems were determined, as before, using the GKLS generator \cite{c43}. When solving the problems, the accuracy $\varepsilon=0.05$ and the reliability $r=5.6$ were used. The results of numerical experiments are presented in Table \ref{tab:4} and Fig.\ref{fig:2}.

@@

This paper proposes an efficient approach for solving computationally intensive multi-objective optimization problems which require computing several Pareto-optimal decisions. It is also assumed that objective functions may be multiextremal, and computation of function values may require a large amount of calculations. The proposed approach is based on the reduction of multi-objective optimization problems to one-dimensional global optimization problems that can be solved using efficient information-statistical algorithms of global search.  The novelty of the proposed approach consists in the possibility of solving several global optimization problems simultaneously, which allows multiple Pareto-optimal decisions to be obtained. Besides, such approach provides for reuse of the computed search information, which considerably reduces computational effort for solving multi-objective optimization problems.
